---
title: "Cell type - disease correlation"
author: "Harm Brugge"
date: "1-11-2016"
output: html_document
---

## Correlation and Wilcoxon-ranked sum test

Here we calculate the Pearson correlation for every cell type and then test for a shift in co-expression for a few disease associated genes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=TRUE)

library(Seurat)
```

## Load data

Load the scRNA seq data and separate the different cell-types.
```{r load_data}

load("./data/pbmc33k.Rda")

TSNEPlot(pbmc33k)

b.cells <- pbmc33k@data[,WhichCells(pbmc33k, c("BCell_A", "BCell_B", "BCell_C"))]
t.h.cells <- pbmc33k@data[,WhichCells(pbmc33k, c("CD4_Naive", "CD4_Memory_Antiviral", "CD4_Memory_TIGIT", "CD4_Memory"))]
t.c.cells <- pbmc33k@data[,WhichCells(pbmc33k, c("CD8Effector_GZMH", "CD8Effector_GZMK", "CD8_Memory", "CD8_Naive"))]
nk.cells <- pbmc33k@data[,WhichCells(pbmc33k, c("NK", "NK_Prss57", "NK_XCL1"))]
monocytes <- pbmc33k@data[,WhichCells(pbmc33k, c("Mono_CD16+", "Mono_CD16+_C1qa", "Mono_Apobec3b", "Mono_Apobec3a", "Mono_CD14+", "Mono_CD14+_Inflam", "Mono_CD14+_Antiviral"))]
megakaryos <- pbmc33k@data[,WhichCells(pbmc33k, "Megakaryocyte")]
dendritic.cells <- pbmc33k@data[,WhichCells(pbmc33k, c("DC_CD141+", "DC_CD1C+"))]

```

Disease genes
```{r}
diabetes.1.genes <- c("AP4B1","BCAR1","C11orf21","C14orf64","C1QTNF6","C7orf71","CCDC101","CCR7","CD226","CDC37P1","CDK2","CLEC16A","CLECL1","CLN3","CSF3","CTLA4","CTSH","DOK6","EIF3C","EIF3CL","ENSG00000232530","ENSG00000241287","ENSG00000256582","ENSG00000257411","ENSG00000258511","ENSG00000261089","ENSG00000264198","ENSG00000266840","ERBB3","FKRP","GDF11","GSDMB","HLA.DMA","HLA.DQA1","HLA.DQB1","HLA.DQB1.AS1","HLA.DQB2","HLA.DRB5","HOTAIRM1","HOXA1","HOXA2","IKZF3","IL27","IL2RA","LIF","MIR320E","NPIPB6","NPIPB7","NPIPB8","NPIPB9","ORMDL3","PRKD2","PTPN2","RAB5B","RABEP2","RPS26","SBK1","SH2B1","SIRPG","SKAP2","STRN4","SULT1A1","SULT1A2","SUOX","TSPAN32","TYK2","ZPBP2")
celiac.genes <- c("AGER","ARHGAP31","BCL9L","C1orf106","CACNA1S","CCDC116","CCR1","CCR2","CCR3","CCRL2","CIITA","CPLX3","CSK","CXCR6","ELMO1","ENSG00000223552","ENSG00000236525","ENSG00000237058","ENSG00000238164","ENSG00000240310","ENSG00000241287","ENSG00000255422","ENSG00000259382","ENSG00000261606","ENSG00000263080","FLT1P1","GPR25","GPSM3","HLA.DMA","HLA.DQB1","HLA.DQB1.AS1","HLA.DRB5","ICOS","ICOSLG","IL18R1","IL1RL1","IRF4","KIAA1109","LMAN1L","LTF","MIR4513","MIR4772","MMEL1","PBX2","PLEK","PTPN2","RMI2","RNF5","THEMIS","TNFRSF14","TTC34","UBASH3A","UBE2L3","ULK3")
ibd.genes <- c("AAGAB","AAMP","ACO2","ACTR1A","ADO","AGAP5","ANKMY1","AP4B1","APEH","ARPC2","ATG16L1","BACH2","BSN","C10orf55","C17orf67","C20orf112","C5orf47","C7orf71","CACNA1S","CALM3","CAMK2G","CARD9","CAST","CCDC116","CCR5","CD226","CD40","CDH3","CEBPB","CELSR3","CISD1","CLN3","COMMD7","CPEB4","CRTC3","CSDC2","CTSW","CUL2","CXCL5","DACT3","DAG1","DENND1B","DESI1","DNLZ","DOCK7","DOK6","EDN3","EFCAB8","EFEMP2","EIF2S2P3","EIF3C","ENSG00000167807","ENSG00000170091","ENSG00000197146","ENSG00000202533","ENSG00000203999","ENSG00000213702","ENSG00000224276","ENSG00000224397","ENSG00000225331","ENSG00000225931","ENSG00000227598","ENSG00000228037","ENSG00000229299","ENSG00000229951","ENSG00000231858","ENSG00000231993","ENSG00000233077","ENSG00000234132","ENSG00000235582","ENSG00000236525","ENSG00000237604","ENSG00000238164","ENSG00000238280","ENSG00000242288","ENSG00000242687","ENSG00000243628","ENSG00000243696","ENSG00000247121","ENSG00000248636","ENSG00000249141","ENSG00000251572","ENSG00000254275","ENSG00000256428","ENSG00000256914","ENSG00000257582","ENSG00000261089","ENSG00000261338","ENSG00000261662","ENSG00000262820","ENSG00000266786","ENSG00000266840","ENSG00000267607","ENSG00000268810","ENSG00000269151","ENSG00000269292","ENSG00000269636","ERAP1","ERAP2","ERGIC1","FADS1","FADS2","FEN1","FIBP","FOXR1","FUT2","GNA12","GNG8","GPBAR1","GPR25","GSDMB","HLA.DRB5","HLA.DRB6","HOTAIRM1","HOXA1","HOXA2","ICAM1","ICAM4","ICAM5","ICOSLG","IL18R1","IL1RL1","IL23R","IL27","IL6ST","INPP5D","IP6K2","IPMK","IRF5","IRGM","ITGAL","ITIH4","ITM2BP1","KIF11","KIR2DL1","KIR3DL1","KIR3DL2","KSR1","LIME1","LNPEP","LRRK2","LY75","LY75.CD302","MAP3K8","MAPRE1","MEI1","MIR1908","MIR3939","MIR4772","MIR611","MMEL1","MRPL20","MRPL23","MST1","MUS81","MXRA8","MYRF","NDFIP1","NFATC1","NFKB1","NKX2.3","NOTCH2","NPIPB6","NPIPB7","NPIPB8","NPIPB9","NTN5","ORMDL3","P4HA2","PARK7","PEBP1P3","PF4","PFKFB4","PLAU","PLCH2","PLCL1","PMM1","POLR3H","PPP5C","PPP5D1","PRKAB1","PRKCB","PSMB8","PSMB9","PTGIR","PTGS2","PTPN2","PTPRC","RASGRP1","RBM22","REG4","RFTN2","RIPK2","RN7SL657P","RNASET2","RNF145","RNFT1","RPL9P29","RPS6KA2","SAG","SCNN1D","SEC1P","SKAP2","SLC22A5","SLC25A20","SLC26A6","SLC2A4RG","SLC35C2","SLC35D1","SLMO2","SMIM3","SMURF1","SNAPC4","SP140","SUFU","SULT1A2","SYNGR1","TEF","TM9SF4","TMBIM1","TMEM180","TMEM258","TNFRSF14","TNFSF8","TOB2","TSPAN14","TYK2","UBE2L3","UBQLN4","USP36","ZBTB38","ZBTB46","ZFP90","ZGLP1","ZGPAT","ZNF300P1","ZNF831","ZPBP2")
rheuma.genes <- c("ACSL6","AFF3","AHNAK2","ALS2CR12","ANKRD55","AP4B1","ARAP1","ARAP1-AS2","ARID5B","ATG5","BCL9L","BLK","CCDC157","CCR6","CD226 ","CD40","CD5","CFLAR","CTLA4","DDX6","ENSG00000228403","ENSG00000230955","ENSG00000237058","ENSG00000238164","ENSG00000254275","ENSG00000254774","ENSG00000255354","ENSG00000255518","ENSG00000256914","ENSG00000262211","ENSG00000264198","ENSG00000266840","ENSG00000267520","ENSG00000269468","ENSG00000269954","ERBB3","FADS1","FADS2","FAM167A","GPSM3","GSDMB","HLA-DQA1","HLA-DQA2","HLA-DRB1","HLA-DRB5","HLA-DRB6","HLA-DRB9","IFNAR1","IL6R","IL6ST","INPP5B","IRF5","JAZF1","KCTD20","MED1","MMEL1","MYRF","NOTCH4","ORMDL3","PADI4","PAM","PBX2","PDE2A","PHF19","PSMD5","RPS26","RTKN2","SHE","SLC35C2","SPRED2","SUOX","SYNGR1","TBC1D10A","TNFRSF14","TRAF1","TTC34","UBASH3A","UTS2","VPS37C","ZPBP2")
```

## B cells

Calculate the correlation, r-squared and totals.
```{r b_cell_correlation}
load("data/cell_type_cor.Rda")

b.cells <- b.cells[apply(b.cells[,-1], 1, function(x) !all(x==0)),] # remove zero rows
b.cells.cor <- cor(t(as.matrix(b.cells)))
b.cells.r2 <- b.cells.cor ^ 2
b.cells.total.expression <- rowSums(as.matrix(b.cells)) # total expression per gene
b.cells.total.r2 <- rowSums(b.cells.r2)
```

Plot the the r-squared against the total expression of all the genes in the b-cells
```{r b_cell_plot}
par(mfrow = c(1, 3))
plot(b.cells.total.r2~b.cells.total.expression, ylab = "Sum of r2", xlab ="Total expression", main = "All genes", cex.lab = 1.5, cex.main = 1.5, pch=20, col=rgb(0,0,0,alpha=0.4), cex=0.5)
plot(b.cells.r2["RPL10",]~b.cells.total.expression, ylim=c(0,0.1), ylab = "R2", xlab ="Total expression", main = "Highly expressed gene", cex.lab = 1.5, cex.main = 1.5, pch=20, col=rgb(0,0,0,alpha=0.4), cex=0.5)
plot(b.cells.r2[1,]~b.cells.total.expression, ylim=c(0,0.1), ylab = "R2", xlab ="Total expression", main = "Lowly expressed gene", cex.lab = 1.5, cex.main = 1.5, pch=20, col=rgb(0,0,0,alpha=0.4), cex=0.5)
par(mfrow = c(1, 1))


```

Select the diabetes and non-diabetes genes from the b-cells, and calculate the correlation and r-squared
```{r b_cells_dia}
diabetes.1.in.b.cells <- rownames(b.cells) %in% diabetes.1.genes

b.cells.not.dia <- b.cells[!diabetes.1.in.b.cells,]
b.cells.not.dia.r2 <- b.cells.r2[!diabetes.1.in.b.cells, !diabetes.1.in.b.cells]

b.cells.dia <- b.cells[diabetes.1.in.b.cells,]
b.cells.dia.r2 <- b.cells.r2[diabetes.1.in.b.cells, diabetes.1.in.b.cells]

b.cells.dia.r2.upper <- b.cells.dia.r2[upper.tri(b.cells.dia.r2)]
b.cells.not.dia.r2.upper <- b.cells.not.dia.r2[upper.tri(b.cells.not.dia.r2)]
```

Explore the b-cells/diabetes data
```{r b_cells_dia_explore}
boxplot(b.cells.dia.r2, ylim=c(0,0.01))

dim(b.cells)

par(mfrow = c(1, 2))
plot(b.cells.total.r2~b.cells.total.expression, ylab = "Sum of r2", xlab ="Total expression", main = "All genes", cex.lab = 1.2, cex.main = 1.5, pch=20, col=rgb(0,0,0,alpha=0.4), cex=0.5)
abline(v = 200, lty = 2, lwd=2, col="red")
plot(rowSums(b.cells.dia.r2)~rowSums(as.matrix(b.cells.dia)), xlim=c(0,10000), ylab = "Sum of r2", xlab ="Total expression", main = "Diabetes genes", cex.lab = 1.2, cex.main = 1.5, pch=20)
abline(v = 200, lty = 2, lwd=2, col="red")
par(mfrow = c(1, 1))

length(b.cells.total.expression[b.cells.total.expression > 200])
length(b.cells.total.expression[b.cells.total.expression < 200])

sd(b.cells.dia.r2)
sd(b.cells.not.dia.r2)

mean(as.vector(b.cells.dia))
mean(as.vector(b.cells.not.dia))

sd(b.cells.dia)
sd(b.cells.not.dia)
```

Wilcoxon test of diabetes gene correlaton against 1*10^6 random non-disease correlation 
```{r}
b.cells.not.dia.r2.upper.random <- sample(b.cells.not.dia.r2.upper, 1e6)
wilcox.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper.random)
b.cells.not.dia.r2.upper.random <- sample(b.cells.not.dia.r2.upper, 1e6)
wilcox.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper.random)
b.cells.not.dia.r2.upper.random <- sample(b.cells.not.dia.r2.upper, 1e6)
wilcox.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper.random)
```


# Iterative wilcoxon test

Iteratively calculate the p-values with the wilcoxon test against random non-disease genes, with the same sample size as the disease genes.
```{r wilcox_itter}
wilcox.n.test <- function(x, y, rep=500, sample.size=length(x)) {
  p.values <- numeric(rep)
  for (i in 1:rep) p.values[i] <- wilcox.test(x, sample(y, sample.size))$p.value
  return(p.values)
}
length(b.cells.dia.r2.upper)
dim(b.cells.dia)

p.values.sample.same <- wilcox.n.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper)
p.values.sample.1e3 <- wilcox.n.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper, rep=500, sample.size = 1e3)
p.values.sample.1e4 <- wilcox.n.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper, rep=500, sample.size = 1e4)
p.values.sample.1e5 <- wilcox.n.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper, rep=500, sample.size = 1e5)
p.values.sample.1e6 <- wilcox.n.test(b.cells.dia.r2.upper, b.cells.not.dia.r2.upper, rep=500, sample.size = 1e6)
mean(p.values)
boxplot(p.values.sample.1000)
boxplot(p.values.sample.1e3)
mean(p.values.sample.1e3)
mean(p.values.sample.1e4)
mean(p.values.sample.1e5)
mean(p.values.sample.1e6)
boxplot(p.values.sample.1e4)
boxplot(p.values.sample.1e5)
boxplot(p.values.sample.1e6)

p.values.matrix <- cbind(p.values.sample.same,p.values.sample.1e3, p.values.sample.1e4, p.values.sample.1e5, p.values.sample.1e6)
colnames(p.values.matrix) <- c("N=41", "N=1e3", "N=1e4", "N=1e5", "N=1e6")

boxplot(p.values.matrix[,5], cex.lab=1.5)

dim(b.cells)[1]*dim(b.cells)[2]
nnzero(b.cells, na.counted=T)
sums <- rowSums(as.matrix(b.cells)==0)
percentages.zero <- sums / dim(b.cells)[2]

percentages.zero[1:20]
sums["MS4A1"]

hist(percentages.zero, xlim = c(0.8, 1), breaks = 400, main = "Occurrence of zero values", xlab = "Zero values")

b.cells[1:5,1:5]

total.nnzero <- nnzero(pbmc33k.merged@scale.data, na.counted=T)
total.counts <- dim(pbmc33k.merged@scale.data)[1]*dim(pbmc33k.merged@scale.data)[2]

pbmc33k.merged@scale.data[1:5,1:5]

total.nnzero/total.counts * 100
```

# T-cells

Calculate the correlation
```{r }
t.h.cells <- t.h.cells[apply(t.h.cells[,-1], 1, function(x) !all(x==0)),]
t.h.cells.cor <- cor(t(as.matrix(t.h.cells)))
t.h.cells.r2 <- t.h.cells.cor ^ 2

diabetes.1.in.t.h.cells <- rownames(t.h.cells) %in% diabetes.1.genes

t.h.cells.dia <- t.h.cells[diabetes.1.in.t.h.cells,]
t.h.cells.dia.r2 <- t.h.cells.r2[diabetes.1.in.t.h.cells,diabetes.1.in.t.h.cells]
t.h.cells.dia.r2.upper <- t.h.cells.dia.r2[upper.tri(t.h.cells.dia.r2)]

t.h.cells.not.dia <- t.h.cells[!diabetes.1.in.t.h.cells,]
t.h.cells.not.dia.r2 <- t.h.cells.r2[!diabetes.1.in.t.h.cells,!diabetes.1.in.t.h.cells]
```


```{r }
plot(rowSums(t.h.cells.dia.r2)~rowSums(as.matrix(t.h.cells.dia)), xlim=c(0,2000))
```


```{r }

```


```{r }

```


```{r }

```



